COFFEE: an Optimizing Compiler\\for Finite Element Local Assembly

### 2014-04-08 paper/bare_conf.tex
Re-factor a few paragraphs

% Padding and Data Alignment
\subsection{Padding and Data Alignment}
Auto-vectorization of assembly code computing the element matrix can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length $vl$. Data alignment is enforced in two steps. Initially, all arrays are allocated to addresses that are multiples of $vl$. Then, two dimensional arrays are padded by rounding the number of columns to the nearest multiple of $vl$. For instance, assume the original size of a basis function array is 3$\times$3 and that the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use 64-bits double-precision floating-point values. In this example, a padded version of the array will have size 3$\times$4. The compiler is explicitly informed about data alignment using a suitable pragma. Padding of all two dimensional arrays involved in the evaluation of the element matrix also allows us to safely round the loop trip count to the nearest multiple of $vl$. This avoids the introduction of a remainder (scalar) loop from the compiler, which would render vectorization less efficient.

@@

\begin{algorithm}[t]
\small
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W3[3] ALIGN = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4] ALIGN;\\
~~~~double LI$\_$1[4] ALIGN;\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+\\
~~~~~~~~~~~~~~~~~~~~+LI$\_$1[k]*LI$\_$1[j])*det*W3[i]);\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code generated by Firedrake when padding, data alignment, and $licm$ are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}. Data alignment and padding are for an AVX machine. In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$, so they can be precomputed once in a single loop $r$; in general, this might not be the case.}
\label{code:helmholtz-licm}
\end{algorithm}

@@

The storage layout of $A$, however, is incorrect after the application of this outer-product-based vectorization ($op$-$vect$, in the following). It can be efficiently restored with a sequence of vector shuffles following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the \texttt{ijk} loop nest. The generated pseudo-code for the simple Helmholtz problem when using $op$-$vect$ is shown in Figure~\ref{code:helmholtz-opvect}.

\begin{figure}[t]
\centerline{\includegraphics[scale=0.7]{Pictures/vect-restore.pdf}}
\caption{Restoring the storage layout after $op$-$vect$. The figure shows how 4$\times$4 elements in the top-left block of the element matrix $A$ can be moved to their correct positions. Each rotation, represented by a group of three same-colored arrows, is implemented by a single vector shuffle intrinsics.}
\label{fig:restore-layout}
\end{figure}

@@

\begin{algorithm}[t]
\small
\KwSty{void} helmholtz(double A[8][8], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$6; i++) $\lbrace$ \\
~~~~// Do loop-invariant code motion \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~~~// Call $Load$ and $set$ intrinsics \\
~~~~~~~~// Compute A[1,1],A[2,2],A[3,3],A[4,4] \\
~~~~~~~~// One $permute\_pd$ intrinsics per $k$-loop $load$\\
~~~~~~~~// Compute A[1,2],A[2,1],A[3,4],A[4,3] \\
~~~~~~~~// One $permute2f128\_pd$ intrinsics per $k$-loop $load$\\
~~~~~~~~// ...\\
~~~~~~$\rbrace$\\
~~~~// Do Remainder loop (from $j=4$ to $j=6$)\\
~~$\rbrace$\\
~~// Restore the storage layout:\\
~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) $\lbrace$\\
~~~~$\_\_$m256d r0, r1, r2, r3, r4, r5, r6, r7;\\
~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~r0 = $\_$mm256$\_$load$\_$pd ($\&$A[j+0][k]);\\
~~~~~~// Load A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~~~r4 = $\_$mm256$\_$unpackhi$\_$pd (r1, r0);\\
~~~~~~r5 = $\_$mm256$\_$unpacklo$\_$pd (r0, r1);\\
~~~~~~r6 = $\_$mm256$\_$unpackhi$\_$pd (r2, r3);\\
~~~~~~r7 = $\_$mm256$\_$unpacklo$\_$pd (r3, r2);\\
~~~~~~r0 = $\_$mm256$\_$permute2f128$\_$pd (r5, r7, 32);\\
~~~~~~r1 = $\_$mm256$\_$permute2f128$\_$pd (r4, r6, 32);\\
~~~~~~r2 = $\_$mm256$\_$permute2f128$\_$pd (r7, r5, 49);\\
~~~~~~r3 = $\_$mm256$\_$permute2f128$\_$pd (r6, r4, 49);\\
~~~~~~$\_$mm256$\_$store$\_$pd ($\&$A[j+0][k], r0);\\
~~~~~~// Store A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~$\rbrace$\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code generated by Firedrake when padding, data alignment, $licm$ and $op$-$vect$ are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}. Here, we assume the polynomial order is $p=2$, since $op$-$vect$ can not be used when an iteration space dimension is smaller than the vector length. The original size of the $j$-$k$ iteration space (i.e. before padding was applied) was 6$\times$6. In this example, the unroll-and-jam factor is 1.}
\label{code:helmholtz-opvect}
\end{algorithm}

@@

\begin{algorithm}[t]
\small
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
%~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
~~~~// Same code as in Figure~\ref{code:helmholtz-licm} up to line 15 \\
%~~// Declaration of basis function matrices \\
%~~\\
%~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
%~~~~double LI$\_$0[4] ALIGN;\\
%~~~~double LI$\_$1[4] ALIGN;\\
%~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
%~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
%~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
%~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j])*det*W3[i];\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
$\rbrace$
\caption{Local assembly code generated by Firedrake when $split$ is applied to the optimized Helmholtz problem given in Figure~\ref{code:helmholtz-licm}. In this example, the split factor is 2.}
\label{code:helmholtz-split}
\end{algorithm}

@@

Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations. At run-time, this high-level specification is translated by the Fenics Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels. ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}. The output of COFFEE is C code, which is eventually provided to PyOP2~\cite{pyop2isc}, where just-in-time compilation and parallel execution over the unstructured mesh take place. Because of the large number of (possibly parametric) transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem. Auto-tuning might be used, although at the moment we avoid it to minimize the run-time overhead (the generation of the AST occurs at run-time as soon as problem-specific data are available). Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.

The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}. Initially, an AST is inspected, looking for the presence of iteration spaces and domain-specific information provided by the higher layer. If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied. This feature, currently under development, has been proved to be useful in several finite volume applications~\cite{inter-kernel-vect}. The second transformation step is applied if the backend is a manycore machine, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{pyop2isc}. Then, an ordered sequence of optimization steps are executed. Application of $licm$ must precede padding and data alignment, due to the introduction of temporary arrays. Based on a cost model, loop interchange, $split$, and $op$-$vect$ may be introduced. Their implementation is based on analysis and transformation of the kernel's AST. When $op$-$vect$ is selected, the compiler outputs proper AVX or LRBNi intrinsics code. Any possible corner cases are handled: for example, if $op$-$vect$ is to be applied, but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.

All loops are interchangeable provided that temporaries are introduced if the nest is not perfect. For the employed storage layout, the loop permutations \texttt{ijk} and \texttt{ikj} are likely to maximize performance. Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration; experiments showed that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop. The choice between \texttt{ijk} and \texttt{ikj} depends on the number of load instructions that can be hoisted out of the innermost dimension. Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers remain available to carry out the computation of the element matrix.

Loop unroll and unroll-and-jam of outer loops are fundamental to expose instruction-level parallelism and register reuse, so tuning critical parameters, like the unroll factor, becomes of great importance. It is our experience (inspection of assembly code, comparison with other hand-made implementations), however, that for assembly kernels, where the loop nest is affine, bounds are usually very small and known at compile-time, and memory accesses are unit-stride, recent versions of a vendor compiler like $intel$'s employ cost models capable of estimating close-to-optimal values for such parameters. We leave therefore the backend compiler in charge to select the unroll factor. This choice also simplifies the COFFEE's cost model. The only situation in which we explicitly unroll-and-jam a loop is when $op$-$vect$ is used, since the transformed code seems to prevent the $intel$ compiler from applying this optimization, even if specific pragmas are added. Note that, regardless of the output of the cost model, the unroll-and-jam factor never exceeds the actual size of the outer loop (line 25 in Figure~\ref{algo:PyOP2Compiler}, assuming the default loop order \texttt{ijk}).

The cost model is shown in Figure~\ref{algo:applyCostModel}. It takes into account the number of available logical vector registers, $n\_regs$, and the number of unique variables accessed: $n\_consts$ counts variables independent of both $j$ and $k$ loops and temporary registers, $n\_outer\_arrays$ counts $j$-dependent variables, and $n\_inner\_arrays$ counts $k$-dependent variables, assuming the \texttt{ijk} loop order. These values are used to estimate unroll-and-jam and split factors for $op$-$vect$ and $split$. If a factor is 0, then the corresponding transformation is not applied. The $split$ transformation is triggered whenever the number of hoistable terms is larger than the available registers along the outer dimension (lines 3-8), which is approximated as half of the total (line 2). A split factor of $n$ means that the assembly expression should be ``cut'' into $n$ sub-expressions. Depending on the structure of the assembly expression, each sub-expression might end up accessing a different number of arrays; the cost model is simplified by assuming that all sub-expressions are of the same size. The unroll-and-jam factor for the $op$-$vect$ transformation is determined as a function of the available logical registers, i.e. those not used for storing hoisted terms (line 9-11). Finally, the profitability of loop interchange is evaluated (line 12-16).

@@

Experiments were run on two Intel machines, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Phi. The \texttt{icc 13.1} compiler was used, with optimization level \texttt{-O2} and with auto-vectorization enabled (\texttt{-xAVX} on the Sandy Bridge, and \texttt{TODO} on the Phi). Other optimization levels performed, in general, slightly worse than \texttt{-O2}. Our code transformations were evaluated in three real-world problems based on the following PDEs:
\begin{itemize}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{itemize}
The code was written in UFL and then executed over real unstructured meshes through Firedrake. The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}. For Advection-Diffusion, the ``Diffusion'' equation, which uses the same differential operators as Helmholtz, is considered. In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$. Burgers is a time-dependent problem, i.e. the assembly is recalculated every time step based on the result of previous iterations. It employs differential operators different from those of Helmholtz, which has a major impact on the generated assembly code (Figure~\ref{code:burgers}), where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are accessed.

These problems were studied varying both the shape of mesh elements and the polynomial order $p$ of the method. Intuitively, the bigger the element shape and $p$, the larger is the iteration space. Triangles (2D), tetrahedron (3D), and prisms (3D) were tested as element shape. For instance, in the case of Helmholtz with $p=1$, the size of the $j$ and $k$ loops for the three kind of elements is, respectively, $3$, $4$, and $6$. Moving from 2D to 3D has the effect of increasing the number of basis function arrays, since, conceptually, the behaviour of the equation has now to be approximated also along a third axis. On the other hand, the polynomial order affects only the problem size (the three loops $i$, $j$, and $k$, and, as a consequence, the size of $X$ and $Y$ arrays). A range of polynomial orders, from $p=1$ to $p=4$, were tested; higher polynomial orders are excluded from the study because of current Firedrake limitations. In such a large space of problems, the size of the element matrix rarely exceeds 30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.

@@

\begin{table*}[t]\normalsize
\massmatrixtwodresultsnorms
\label{table:perf-results-licm}
\caption{Impact of generalized loop-invariant code motion ($licm$ column) on the Helmholtz, Diffusion and Burgers problems, for three kind of elements belonging to the Lagrange family (triangle, tetrahedron, prism), for the range of polynomial orders $p \in \lbrace$1, 4$\rbrace$. Each entry indicates the range of speed-ups obtained over the non-optimized implementation. The column $licm$-$ap$ illustrates the combination of $licm$ with data alignment and padding. Results are shown for both the Sandy Bridge and the Phi machine.}
\end{table*}

>>>>

% Padding and Data Alignment
\subsection{Prerequisites for Effective Auto-vectorization: Padding and Data Alignment}
Auto-vectorization of assembly code computing the element matrix can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length $vl$. Data alignment is enforced in two steps. Initially, all arrays are allocated to addresses that are multiples of $vl$. Then, two dimensional arrays are padded by rounding the number of columns to the nearest multiple of $vl$. For instance, assume the original size of a basis function array is 3$\times$3 and that the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use 64-bits double-precision floating-point values. In this example, a padded version of the array will have size 3$\times$4. The compiler is explicitly informed about data alignment using a suitable pragma. Padding of all two dimensional arrays involved in the evaluation of the element matrix also allows us to safely round the loop trip count to the nearest multiple of $vl$. This avoids the introduction of a remainder (scalar) loop from the compiler, which would render vectorization less efficient.

@@

\begin{algorithm}[t]
\small
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W3[3] ALIGN = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4] ALIGN;\\
~~~~double LI$\_$1[4] ALIGN;\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+\\
~~~~~~~~~~~~~~~~~~~~+LI$\_$1[k]*LI$\_$1[j])*det*W3[i]);\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code generated by Firedrake when padding, data alignment, and \emph{licm} are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}. Data alignment and padding are for an AVX machine. In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$, so they can be precomputed once in a single loop $r$; in general, this might not be the case.}
\label{code:helmholtz-licm}
\end{algorithm}

@@

The storage layout of $A$, however, is incorrect after the application of this outer-product-based vectorization (\emph{op-vect}, in the following). It can be efficiently restored with a sequence of vector shuffles following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the \texttt{ijk} loop nest. The generated pseudo-code for the simple Helmholtz problem when using \emph{op-vect} is shown in Figure~\ref{code:helmholtz-opvect}.

\begin{figure}[t]
\centerline{\includegraphics[scale=0.7]{Pictures/vect-restore.pdf}}
\caption{Restoring the storage layout after \emph{op-vect}. The figure shows how 4$\times$4 elements in the top-left block of the element matrix $A$ can be moved to their correct positions. Each rotation, represented by a group of three same-colored arrows, is implemented by a single vector shuffle intrinsics.}
\label{fig:restore-layout}
\end{figure}

@@

\begin{algorithm}[t]
\small
\KwSty{void} helmholtz(double A[8][8], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$6; i++) $\lbrace$ \\
~~~~// Do loop-invariant code motion \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~~~// Call $Load$ and $set$ intrinsics \\
~~~~~~~~// Compute A[1,1],A[2,2],A[3,3],A[4,4] \\
~~~~~~~~// One $permute\_pd$ intrinsics per $k$-loop $load$\\
~~~~~~~~// Compute A[1,2],A[2,1],A[3,4],A[4,3] \\
~~~~~~~~// One $permute2f128\_pd$ intrinsics per $k$-loop $load$\\
~~~~~~~~// ...\\
~~~~~~$\rbrace$\\
~~~~// Do Remainder loop (from $j=4$ to $j=6$)\\
~~$\rbrace$\\
~~// Restore the storage layout:\\
~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) $\lbrace$\\
~~~~$\_\_$m256d r0, r1, r2, r3, r4, r5, r6, r7;\\
~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~r0 = $\_$mm256$\_$load$\_$pd ($\&$A[j+0][k]);\\
~~~~~~// Load A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~~~r4 = $\_$mm256$\_$unpackhi$\_$pd (r1, r0);\\
~~~~~~r5 = $\_$mm256$\_$unpacklo$\_$pd (r0, r1);\\
~~~~~~r6 = $\_$mm256$\_$unpackhi$\_$pd (r2, r3);\\
~~~~~~r7 = $\_$mm256$\_$unpacklo$\_$pd (r3, r2);\\
~~~~~~r0 = $\_$mm256$\_$permute2f128$\_$pd (r5, r7, 32);\\
~~~~~~r1 = $\_$mm256$\_$permute2f128$\_$pd (r4, r6, 32);\\
~~~~~~r2 = $\_$mm256$\_$permute2f128$\_$pd (r7, r5, 49);\\
~~~~~~r3 = $\_$mm256$\_$permute2f128$\_$pd (r6, r4, 49);\\
~~~~~~$\_$mm256$\_$store$\_$pd ($\&$A[j+0][k], r0);\\
~~~~~~// Store A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~$\rbrace$\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code generated by Firedrake when padding, data alignment, \emph{licm} and \emph{op-vect} are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}. Here, we assume the polynomial order is $p=2$, since \emph{op-vect} can not be used when an iteration space dimension is smaller than the vector length. The original size of the $j$-$k$ iteration space (i.e. before padding was applied) was 6$\times$6. In this example, the unroll-and-jam factor is 1.}
\label{code:helmholtz-opvect}
\end{algorithm}

@@

\begin{algorithm}[t]
\small
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
%~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
~~~~// Same code as in Figure~\ref{code:helmholtz-licm} up to line 15 \\
%~~// Declaration of basis function matrices \\
%~~\\
%~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
%~~~~double LI$\_$0[4] ALIGN;\\
%~~~~double LI$\_$1[4] ALIGN;\\
%~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
%~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
%~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
%~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j])*det*W3[i];\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
$\rbrace$
\caption{Local assembly code generated by Firedrake when \emph{split} is applied to the optimized Helmholtz problem given in Figure~\ref{code:helmholtz-licm}. In this example, the split factor is 2.}
\label{code:helmholtz-split}
\end{algorithm}

@@

Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations. At run-time, the high-level specification is translated by the Fenics Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels. ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}. The output of COFFEE, C code, is eventually provided to PyOP2~\cite{pyop2isc}, where just-in-time compilation and execution over the discretized domain take place. Because of the large number of (possibly parametric) transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem. Auto-tuning might be used, although it would significantly increase the run-time overhead, since the generation of ASTs occurs at run-time as soon as problem-specific data are available. Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.

The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}. Initially, an AST is inspected, looking for the presence of iteration spaces and domain-specific information provided by the higher layer. If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied. This feature, currently under development, has been proved to be useful in several finite volume applications~\cite{inter-kernel-vect}. The second transformation step is applied if the backend is a manycore architecture, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{pyop2isc}. Then, an ordered sequence of optimization steps are executed. Application of \emph{licm} must precede padding and data alignment, due to the introduction of temporary arrays. Based on a cost model, loop interchange, \emph{split}, and \emph{op-vect} may be introduced. Their implementation is based on analysis and transformation of the kernel's AST. When \emph{op-vect} is selected, the compiler outputs proper AVX or LRBNi intrinsics code. Any possible corner cases are handled: for example, if \emph{op-vect} is to be applied, but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.

All loops are interchangeable provided that temporaries are introduced if the nest is not perfect. For the employed storage layout, the loop permutations \texttt{ijk} and \texttt{ikj} are likely to maximize performance. Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required for every iteration. Experiments showed that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop. The choice between \texttt{ijk} and \texttt{ikj} depends on the number of load instructions that can be hoisted out of the innermost dimension. Our compiler chooses the loop along which the number of invariant loads is smaller as outer so that more registers remain available to carry out the computation of the element matrix.

Loop unroll and unroll-and-jam of outer loops are fundamental to expose instruction-level parallelism, so tuning critical parameters, like the unroll factor, is of great importance. However, from our experience of inspecting assembly code and comparing with other hand-written implementations, recent versions of a vendor compiler like Intel's employ cost models capable of estimating close-to-optimal values for such parameters. This is particularly true for assembly kernels, where the loop nest is affine, bounds are usually very small and known at compile-time, and memory accesses are unit-stride. We therefore leave the backend compiler in charge of selecting the unroll factor. This choice also simplifies the COFFEE's cost model. The only situation in which we explicitly unroll-and-jam a loop is when \emph{op-vect} is used, since the transformed code seems to prevent the Intel compiler from applying this optimization, even if specific pragmas are added. Note that, regardless of the output of the cost model, the unroll-and-jam factor never exceeds the actual size of the outer loop (line 25 in Figure~\ref{algo:PyOP2Compiler}, assuming the default loop order \texttt{ijk}).

The cost model is shown in Figure~\ref{algo:applyCostModel}. It takes into account the number of available logical vector registers, $n\_regs$, and the number of unique variables accessed: $n\_consts$ counts variables independent of both $j$ and $k$ loops and temporary registers, $n\_outer\_arrays$ counts $j$-dependent variables, and $n\_inner\_arrays$ counts $k$-dependent variables, assuming the \texttt{ijk} loop order. These values are used to estimate unroll-and-jam and split factors for \emph{op-vect} and \emph{split}. If a factor is 0, then the corresponding transformation is not applied. The \emph{split} transformation is triggered whenever the number of hoistable terms is larger than the available registers along the outer dimension (lines 3-8), which is approximated as half of the total (line 2). A split factor of $n$ means that the assembly expression should be ``cut'' into $n$ sub-expressions. Depending on the structure of the assembly expression, each sub-expression might end up accessing a different number of arrays; the cost model is simplified by assuming that all sub-expressions are of the same size. The unroll-and-jam factor for the \emph{op-vect} transformation is determined as a function of the available logical registers, i.e. those not used for storing hoisted terms (line 9-11). Finally, the profitability of loop interchange is evaluated (line 12-16).

@@

Experiments were run on two Intel machines, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Xeon Phi. The \texttt{icc 13.1} compiler was used, with optimization level \texttt{-O2} and with auto-vectorization enabled (\texttt{-xAVX} on the Sandy Bridge, and \texttt{TODO} on the Phi). Other optimization levels performed, in general, slightly worse than \texttt{-O2}. Our code transformations were evaluated in three real-world problems based on the following PDEs:
\begin{itemize}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{itemize}
The code was written in UFL and then executed over real unstructured meshes through Firedrake. The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}. For Advection-Diffusion, the ``Diffusion'' equation, which uses the same differential operators as Helmholtz, is considered. In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$. Burgers is a non-linear problem employing differential operators different from those of Helmholtz, which has a major impact on the generated assembly code (Figure~\ref{code:burgers}), where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are accessed.

These problems were studied varying both the shape of mesh elements and the polynomial order $p$ of the method, whereas the element family, Lagrange, is fixed. Intuitively, the bigger the element shape and $p$, the larger is the iteration space. Triangles (2D), tetrahedron (3D), and prisms (3D) were tested as element shape. For instance, in the case of Helmholtz with $p=1$, the size of the $j$ and $k$ loops for the three kind of elements is, respectively, $3$, $4$, and $6$. Moving from 2D to 3D has the effect of increasing the number of basis function arrays, since, conceptually, the behaviour of the equation has now to be approximated also along a third axis. On the other hand, the polynomial order affects only the problem size (the three loops $i$, $j$, and $k$, and, as a consequence, the size of $X$ and $Y$ arrays). A range of polynomial orders, from $p=1$ to $p=4$, were tested; higher polynomial orders are excluded from the study because of current Firedrake limitations. In such a large space of problems, the size of the element matrix rarely exceeds 30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.

@@

\begin{table*}[t]\normalsize
\massmatrixtwodresultsnorms
\label{table:perf-results-licm}
\caption{Impact of generalized loop-invariant code motion (\emph{licm} column) on the Helmholtz, Diffusion and Burgers problems, for three kind of elements belonging to the Lagrange family (triangle, tetrahedron, prism), for the range of polynomial orders $p \in \lbrace$1, 4$\rbrace$. Each entry indicates the range of speed-ups obtained over the non-optimized implementation. The column \emph{licm-ap} illustrates the combination of \emph{licm} with data alignment and padding. Results are shown for both the Sandy Bridge and the Xeon Phi architectures.}
\end{table*}

Table~\ref{table:perf-results-licm} illustrates the speed-ups obtained on the Sandy Bridge and the Phi machines when \emph{licm}, data alignment, and padding are used, over non-transformed code. We distinguish between \emph{licm} and \emph{licm-ap}; the latter makes use of padding and data alignment. Inspection of assembly code generated by \texttt{icc} confirmed all limitations described in Section~\ref{sec:code-transf-licm}: only sub-expressions invariant with respect to outer loops are hoisted and, interestingly, not vectorized. This motivates the usually significant gain obtained. Padding and data alignment enhance, in general, the quality of SIMD auto-vectorization. Sometimes the run-time of \emph{licm-ap} is similar to that of \emph{licm} because the element matrix size is, without padding, already a multiple of the vector length, and data is automatically aligned. Occasionally \emph{licm-ap} is slower than \emph{licm} (e.g. in Burgers 3D p3). This is due to the large number of aligned temporaries introduced by \emph{licm}, which probably induces cache associativity conflicts.
